---
title: "Adding metadata to environmental sampling observations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding metadata to environmental sampling observations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `es` package contains a number of functions that will add important metadata (e.g. climate variables, elevation, topography, etc) to your environmental sampling observations. This assumes that you have constructed a data frame that matches the [`template_es_data`](https://gilesjohnr.github.io/es/reference/template_es_data.html) data object. A detailed description of the template data can be found [HERE]().

To retrieve metadata for the location and dates in your data frame, the functions listed in the table below will download them from open source APIs using wrappers of functions from other R packages. These methods should generalized to most locations and dates assuming that the data are available (e.g. very recent dates may not be uploaded to climate APIs yet). The list of metadata functions are:

```{r, echo=FALSE}
x <- data.frame(wrapper_func=as.character(),
                description=as.character(),
                wrapped_pkg=as.character(),
                wrapped_func=as.character(),
                api=as.character())

# get_precip_data

x[1, 'wrapper_func'] <- "[get_precip_data](https://gilesjohnr.github.io/es/reference/get_precip_data.html)"
x[1, 'description'] <- "Cumulative daily millimeters of rainfall"
x[1, 'wrapped_pkg'] <- "[openmeteo](https://cran.r-project.org/web/packages/openmeteo/index.html)"
x[1, 'wrapped_func'] <- "[weather_history](https://rdrr.io/cran/openmeteo/man/weather_history.html)" 
x[1, 'api'] <- "[Open-Meteo Historical Weather API](https://open-meteo.com/en/docs/historical-weather-api)"

# get_temp_data

x[2, 'wrapper_func'] <- "[get_temp_data](https://gilesjohnr.github.io/es/reference/get_temp_data.html)"
x[2, 'description'] <- "[Accumulated Thermal Units (ATUs)](https://en.wikipedia.org/wiki/Accumulated_thermal_unit)"
x[2, 'wrapped_pkg'] <- "[openmeteo](https://cran.r-project.org/web/packages/openmeteo/index.html)"
x[2, 'wrapped_func'] <- "[weather_history](https://rdrr.io/cran/openmeteo/man/weather_history.html)" 
x[2, 'api'] <- "[Open-Meteo Historical Weather API](https://open-meteo.com/en/docs/historical-weather-api)"

# get_esi_data

x[3, 'wrapper_func'] <- "[get_esi_data](https://gilesjohnr.github.io/es/reference/get_esi_data.html)"
x[3, 'description'] <- "[Evapoartive Stress Index (ESI)](https://climateserv.readthedocs.io/en/latest/user/datasets.html#evaporative-stress-index-esi)"
x[3, 'wrapped_pkg'] <- "[chirps](https://docs.ropensci.org/chirps/)"
x[3, 'wrapped_func'] <- "[get_esi](https://docs.ropensci.org/chirps/reference/get_esi.html)" 
x[3, 'api'] <- "[SERVIR ClimateSERV API](https://servirglobal.net/services/climateserv)"

# get_river_discharge_data

x[4, 'wrapper_func'] <- "[get_river_discharge_data](https://gilesjohnr.github.io/es/reference/get_river_discharge_data.html)"
x[4, 'description'] <- "Daily river discharge of the nearest river in cubic meters per second ($m^3/s$)"
x[4, 'wrapped_pkg'] <- "[openmeteo](https://cran.r-project.org/web/packages/openmeteo/index.html)"
x[4, 'wrapped_func'] <- "[river_discharge](https://rdrr.io/cran/openmeteo/man/river_discharge.html)" 
x[4, 'api'] <- "[Open-Meteo Global Flood API](https://open-meteo.com/en/docs/flood-api)"

# get_elevation_data

x[5, 'wrapper_func'] <- "[download_elevation_data](https://gilesjohnr.github.io/es/reference/download_elevation_data.html), [get_elevation_data](https://gilesjohnr.github.io/es/reference/get_elevation_data.html)"
x[5, 'description'] <- "Digital Elevation Model (DEM) giving the height in meters above sea level, spatial resolution is ~100m"
x[5, 'wrapped_pkg'] <- "[elevatr](https://cran.r-project.org/web/packages/elevatr/index.html)"
x[5, 'wrapped_func'] <- "[get_elev_raster](https://www.rdocumentation.org/packages/elevatr/versions/0.99.0/topics/get_elev_raster), [get_elev_point](https://www.rdocumentation.org/packages/elevatr/versions/0.99.0/topics/get_elev_point)"
x[5, 'api'] <- "[Amazon Web Services Terrain Tiles](https://registry.opendata.aws/terrain-tiles/)"

# get_hydro_data

x[6, 'wrapper_func'] <- "[get_hydro_data](https://gilesjohnr.github.io/es/reference/get_hydro_data.html)"
x[6, 'description'] <- "slope, aspect, TWI, flow accumulation, distance to streams"
x[6, 'wrapped_pkg'] <- "[elevatr](https://cran.r-project.org/web/packages/elevatr/index.html), [WhiteboxTools R frontend](https://www.whiteboxgeo.com/wbt-frontends/) "
x[6, 'wrapped_func'] <- "[get_elev_raster](https://www.rdocumentation.org/packages/elevatr/versions/0.99.0/topics/get_elev_raster), [wbt_d_inf_flow_accumulation](https://rdrr.io/rforge/whitebox/man/wbt_d_inf_flow_accumulation.html), [wbt_wetness_index](https://rdrr.io/cran/whitebox/man/wbt_wetness_index.html)"
x[6, 'api'] <- "DEM from [Amazon Web Services Terrain Tiles](https://registry.opendata.aws/terrain-tiles/)"

# get_admin_data

x[7, 'wrapper_func'] <- "[download_admin_data](https://gilesjohnr.github.io/es/reference/download_admin_data.html), [get_admin_data](https://gilesjohnr.github.io/es/reference/get_admin_data.html)"
x[7, 'description'] <- "Names of all administrative areas for each sampling site"
x[7, 'wrapped_pkg'] <- "NA"
x[7, 'wrapped_func'] <- "NA"
x[7, 'api'] <- "GeoJSON files from the [geoBoundaries API](https://www.geoboundaries.org/) (saved as ESRI Shapefiles)"

# download_worldpop_data

x[8, 'wrapper_func'] <- "[download_worldpop_data](https://gilesjohnr.github.io/es/reference/download_worldpop_data.html)"
x[8, 'description'] <- "WorldPop population raster data"
x[8, 'wrapped_pkg'] <- "NA"
x[8, 'wrapped_func'] <- "NA"
x[8, 'api'] <- "GeoTIFF files from the [WorldPop FTP server](https://hub.worldpop.org/geodata/listing?id=29)"

# get_population_catchment

x[9, 'wrapper_func'] <- "[get_population_catchment](https://gilesjohnr.github.io/es/reference/get_population_catchment.html)"
x[9, 'description'] <- "Area of drainage catchments (sq km) and total population within each"
x[9, 'wrapped_pkg'] <- "[WhiteboxTools R frontend](https://www.whiteboxgeo.com/wbt-frontends/) "
x[9, 'wrapped_func'] <- "[wbt_d_inf_flow_accumulation](https://rdrr.io/rforge/whitebox/man/wbt_d_inf_flow_accumulation.html), [wbt_watershed](https://rdrr.io/cran/whitebox/man/wbt_watershed.html)"
x[9, 'api'] <- "User supplied elevation and population rasters."

# get_population_radius

x[10, 'wrapper_func'] <- "[get_population_radius](https://gilesjohnr.github.io/es/reference/get_population_radius.html)"
x[10, 'description'] <- "Total population within a given radius"
x[10, 'wrapped_pkg'] <- "[raster](https://cran.r-project.org/web/packages/raster/index.html), [exactextractr](https://cran.r-project.org/web/packages/exactextractr/index.html) "
x[10, 'wrapped_func'] <- "[buffer](https://rdrr.io/rforge/whitebox/man/wbt_d_inf_flow_accumulation.html), [exactextractr](https://rdrr.io/cran/exactextractr/man/exact_extract.html)"
x[10, 'api'] <- "User supplied population raster."

knitr::kable(
     x,
     col.names = c("`es` wrapper", 
                   "Description", 
                   "Package", 
                   "Function", 
                   "API"),
     table.attr = "style='width:100%;'"
)


```

