---
title: "Templates for input data sets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To use the data and methods provided in the `es` R package, input data must match the formatting shown in the `template_es_data` and `standard_curves` data objects. These objects are installed along with the `es` package and provide a starting point to calculate derivative quantities, adding spatial metadata, and launching the visualization application. Both template data sources are described in more detail below:

## Template environmental sampling data

The primary data that is required for the methods in the `es` package is shown in the `template_es_data` object. These data have 6 columns and give the time and location of each observation and a Cycle Threshold (Ct) value from a qPCR assay. The code below shows how to view the data template:

```{r, eval=FALSE}
library(es)
head(template_es_data)
date location_id  lat   lon target_name ct_value
1 2020-03-07           1 23.8 90.37    target_0       NA
2 2020-03-07           1 23.8 90.37    target_0       NA
3 2020-03-07           1 23.8 90.37    target_0       NA
4 2020-03-07           1 23.8 90.37    target_0 29.94516
5 2020-03-07           1 23.8 90.37    target_1 31.61178
6 2020-03-07           1 23.8 90.37    target_1 32.22351

str(template_es_data)
'data.frame':	5200 obs. of  6 variables:
     $ date       : IDate, format: "2020-03-07" "2020-03-07" ...
$ location_id: int  1 1 1 1 1 1 1 1 1 1 ...
$ lat        : num  23.8 23.8 23.8 23.8 23.8 23.8 23.8 23.8 23.8 23.8 ...
$ lon        : num  90.4 90.4 90.4 90.4 90.4 ...
$ target_name: chr  "target_0" "target_0" "target_0" "target_0" ...
$ ct_value   : num  NA NA NA 29.9 31.6 ...
```

More detailed descriptions of each variable in the `template_es_data` object are shown in the table below:

```{r, echo=FALSE}
x <- data.frame(variable=as.character(),
                class=as.character(),
                description=as.character())

x[1, 'variable'] <- "date"
x[1, 'class'] <- "Date, IDate"
x[1, 'description'] <- "The date the environmental sample was collected. Format is YYY-MM-DD."

x[2, 'variable'] <- "location_id"
x[2, 'class'] <- "Integer, Character"
x[2, 'description'] <- "A unique identifier for each sampling location."

x[3, 'variable'] <- "lat"
x[3, 'class'] <- "Numeric"
x[3, 'description'] <- "The lattitude of the sampling location in Decimal Degrees (DD)"

x[4, 'variable'] <- "lon"
x[4, 'class'] <- "Numeric"
x[4, 'description'] <- "The longitude of the sampling location in Decimal Degrees (DD)"

x[5, 'variable'] <- "target_name"
x[5, 'class'] <- "Character"
x[5, 'description'] <- "The unique name of each gene target in qPCR assays"

x[6, 'variable'] <- "ct_value"
x[6, 'class'] <- "Numeric"
x[6, 'description'] <- "The Cycle Threshold (Ct) value returned by qPCR assays"

knitr::kable(
     x,
     col.names = c("Variable", 
                   "Class", 
                   "Description")
)
```

## Template standard curve data

Standard curve data provide the results from standardized qPCR assays which relate Ct values to the number of gene copies for a particular target. With these data, the `calc_n_copies()` function will infer the number of gene copies for each of the observed Ct values in a data set formatted according to the `template_es_data` object. To view the template standard curve data, see the code below:

```{r, eval=FALSE}
library(es)
head(template_standard_curve)
  target_name n_copies ct_value
1    target_1    1e+01 31.29322
2    target_1    1e+02 27.73392
3    target_1    1e+03 23.48097
4    target_1    1e+04 18.91412
5    target_1    1e+05 16.68971
6    target_2    1e+01 32.34237

str(template_standard_curve)
'data.frame':	15 obs. of  3 variables:
 $ target_name: chr  "target_1" "target_1" "target_1" "target_1" ...
 $ n_copies   : num  1e+01 1e+02 1e+03 1e+04 1e+05 1e+01 1e+02 1e+03 1e+04 1e+05 ...
 $ ct_value   : num  31.3 27.7 23.5 18.9 16.7 ...
```


More detailed descriptions of each variable in the `template_standard_curve` object are shown in the table below:

```{r, echo=FALSE}
x <- data.frame(variable=as.character(),
                class=as.character(),
                description=as.character())

x[1, 'variable'] <- "target_name"
x[1, 'class'] <- "Character"
x[1, 'description'] <- "The unique name of each gene target in qPCR assays"

x[2, 'variable'] <- "n_copies"
x[2, 'class'] <- "Numeric"
x[2, 'description'] <- "The known number of gene copies for the observation in the standardized qPCR assay"

x[3, 'variable'] <- "ct_value"
x[3, 'class'] <- "Numeric"
x[3, 'description'] <- "The Cycle Threshold (Ct) value returned by qPCR assays"

knitr::kable(
     x,
     col.names = c("Variable", 
                   "Class", 
                   "Description")
)
```


